ftprime: Recording the pedigree: efficient simulation of whole genomes

### 2017-03-05 tssim.tex
simupop simulation script

For a set of coalescence records to be consistent,
it is required that:
\begin{enumerate}
    \item The birth time of any \texttt{children} must be strictly after the birth time of the \texttt{parent}.
    \item The set of intervals on which individual $a$ is a child must be disjoint, for every $a$.
    \item The set of intervals on which individual $a$ is a parent must be disjoint, for every $a$.
\end{enumerate}
These ensure that the coalescence records unambiguously specify a subset of the genealogy at any location on the genome,
but do not ensure that the entire history of the samples is specified.
Furthermore, to allow for efficient algorithms, msprime also requires that
\begin{enumerate} \setcounter{enumi}{3}
    \item The samples must be numbered \texttt{0,...,n-1}, and the smallest non-sampled label must be an internal node.
        \plr{perhaps remove this one}
    \item The list of \texttt{children} in a coalescence record must be sorted.
    \item Records must be sorted in nondecreasing order by birth time of the \texttt{parent}.
    \item Birth times must be strictly greater than zero.
\end{enumerate}

@@

Since we do not know from which ancestors
the eventual genetic samples will inherit,
to record the history of the samples
we must record the results of all reproduction events -- 
the entire ARG.
To do this, and conform with the format above, we maintain
an ordered dictionary \texttt{arg},
with one entry for each parent,
whose entries are of the form
\texttt{id:(time,records)},
where \texttt{id} is the integer ID of the parent,
\texttt{time} is the birth time of the parent,
and \texttt{records} is a list of nonoverlapping coalescence records, sorted by left endpoint.
We update this structure each time a new individual is born.
Suppose that individual $I$ is born at time $t$ to parents $p_0$ and $p_1$,
with recombination breakpoints $0=x_0 < x_1 < x_2 < \cdots < x_n \le 1$,
so that $I$ inherits the genomic segment $[0,x_1)$ from $p_0$,
the segment $[x_1,x_2)$ from $p_1$, and so forth (here $n \ge 1$).
When this occurs, we:
\begin{enumerate}
    \item append a new entry \texttt{I:(t,[])} to the list \texttt{arg}; and
    \item for each $1 \le k \le n$, merge the new record \texttt{(x[k-1],x[k],p[k mod 2],(I,))}
        into the existing records in `arg[p[k mod 2]]`.
\end{enumerate}
We need to merge these records
so as to retain sortedness described above.

>>>>

For a set of coalescence records to be consistent,
it is required that:
\begin{enumerate}
    \item The birth time of any \texttt{children} must be strictly after the birth time of the \texttt{parent}.
    \item The set of intervals on which individual \texttt{a} is a child must be disjoint, for every \texttt{a}.
    \item The set of intervals on which individual \texttt{a} is a parent must be disjoint, for every \texttt{a}.
\end{enumerate}
These ensure that the coalescence records unambiguously specify a subset of the genealogy at any location on the genome,
but do not ensure that the entire history of the samples is specified.
Furthermore, to allow for efficient algorithms, msprime also requires that
\begin{enumerate} \setcounter{enumi}{3}
    \item The samples must be numbered \texttt{0,...,n-1}, and the smallest non-sampled label must be an internal node.
        \plr{perhaps remove this one}
    \item The list of \texttt{children} in a coalescence record must be sorted.
    \item Records must be sorted in nondecreasing order by birth time of the \texttt{parent}.
    \item Birth times must be strictly greater than zero.
\end{enumerate}

@@

Simulation software often gives a unique identifier to each \emph{diploid} individual,
but the natural exchangeable unit of consideration for the ARG (and msprime) is gamete --
so, each diploid individual is the union of two haploid individuals: 
one consisting of the set of chromosomes inherited maternally, the other paternally.
To translate between the two,
we adopt the convention that the maternally inherited set of chromosomes of diploid individual $k$
is given haploid ID $2k$, and the paternally inherited set $2k+1$.
ID numbers referred to below are haploid IDs.

Since we do not know from which ancestors
the eventual genetic samples will inherit,
to record the history of the samples
we must record the results of all reproduction events -- 
the entire ARG.
To do this, and conform with the format above, we maintain
an ordered dictionary \texttt{arg},
with one entry for each parent,
whose entries are of the form
\texttt{id:(time,records)},
where \texttt{id} is the integer ID of the parent,
\texttt{time} is the birth time of the parent,
and \texttt{records} is a list of nonoverlapping coalescence records, sorted by left endpoint.
We update this structure each time a new individual is born.
Suppose that individual $I$ is born at time $t$ to parents $p_0$ and $p_1$,
with recombination breakpoints $0=x_0 < x_1 < x_2 < \cdots < x_n \le L$,
so that $I$ inherits the genomic segment $[0,x_1)$ from $p_0$,
the segment $[x_1,x_2)$ from $p_1$, and so forth (here $n \ge 1$),
and where $L$ is the total length of the chromosome.
When this occurs, we:
\begin{enumerate}
    \item append a new entry \texttt{I:(t,[])} to the list \texttt{arg}; and
    \item for each $1 \le k \le n$, \texttt{merge} the new record \texttt{(x[k-1],x[k],p[k mod 2],(I,))}
        into the existing records in `arg[p[k mod 2]]`.
\end{enumerate}
The second step requires we merge the new record with existing ones
so as to retain the criteria described above.
To \texttt{merge} a record saying that individual \texttt{I} also inherits from this \texttt{parent} on interval \texttt{[a,b)}
into an existing set of records of the form \texttt{(left,right,parent,children)},
we need to 
\begin{enumerate}
    \item split any record for which \texttt{left<a<right} or \texttt{left<b<right} into two at the corresponding endpoint, 
    \item and then add \texttt{I} to \texttt{children} for any intervals overlapping \texttt{[a,b)}.
\end{enumerate}
This is done easily, maintaining the fact that records are not overlapping and are maintained in sorted order.
\plr{could alternatively write out algorithm but it is a bit nitpicky}

% To do this, to \texttt{merge} a record \texttt{(new.left,new.right,parent,new.children)}
% into an existing sorted list of \texttt{records},
% we do the following.
% \begin{enumerate} \setcounter{enumi}{-1}
%     \item Initialize \texttt{here=new.left} and \texttt{k=0}.
%     \item While \texttt{here<right}, let the next entry of \texttt{records} 
%         (ordered by left endpoint) be \texttt{(left,right,parent,children)}.
%         \begin{enumerate}
%             \item If \texttt{right<here}, go to the next record.
%             \item If \texttt{here<left}: 
%                 \begin{enumerate}
%                     \item insert a new record \texttt{(here,min(new.right,left),parent,new.children)} before the current record
%                     \item and set \texttt{here=min(new.right,left)}.
%                 \end{enumerate}
%             \item otherwise, if  TOO MANY STEPS
%         \end{enumerate}
% \end{enumerate}
